---
title: "Modelling Norms for Speeded Tests and Count Data with the Conway-Maxwell-Poisson (CMP) Distribution"
author: "Wolfgang Lenhard & Alexandra Lenhard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling Norms for Speeded Tests and Count Data with the Conway-Maxwell-Poisson (CMP) Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 60),
  collapse = TRUE,
  comment = "#>"
)
options(width = 60)

library(cNORM)
```

# Introduction

In psychological and educational assessment, **speeded tests** constitute a distinct and widely used measurement paradigm. Typical examples include processing speed tasks (e.g., symbol search, coding, digit span in Wechsler batteries), cancellation and attention tests (e.g., test d2), reading fluency tasks, and rapid automatized naming (RAN). In these assessments, items are uniformly easy and the raw score is defined as the **number of correctly processed units within a fixed time limit**.

From a statistical standpoint, these scores represent **open-ended count data** ($y \in \{0, 1, 2, \dots\}$). Conventional count regression models, however, are poorly suited for norming speeded tests:
- **The Poisson distribution** imposes the rigid restriction of *equi-dispersion*, forcing the variance to equal the mean.
- **The Negative Binomial distribution** is strictly restricted to *over-dispersion*.

In speeded cognitive tests, participants typically work at a stable, homogeneous pacing under time constraints, resulting in performance variance that is substantially smaller than the mean — a phenomenon known as **under-dispersion**. 

To resolve this limitation, `cNORM` introduces the **Conway-Maxwell-Poisson (CMP) continuous norming model** (`cnorm.cmp()`). The CMP distribution generalizes the Poisson distribution by introducing a continuous dispersion parameter $\nu$, naturally accommodating over-, equi- and under-dispersion by modelling the dispersion parameter $\nu$.


# Mathematical Foundation

The classical Conway-Maxwell-Poisson distribution (Conway & Maxwell, 1962; Shmueli et al., 2005) is an exponential family distribution with probability mass function (PMF):

$$P(Y = y \mid \mu, \nu) = \frac{1}{Z(\mu, \nu)} \left(\frac{\mu^y}{y!}\right)^\nu, \quad y \in \{0, 1, 2, \dots\}$$

where $\mu > 0$ is the location parameter, $\nu > 0$ is the dispersion parameter, and $Z(\mu, \nu)$ is the normalizing constant:

$$Z(\mu, \nu) = \sum_{j=0}^{\infty} \left(\frac{\mu^j}{j!}\right)^\nu$$


Both the location $\mu(a)$ and dispersion $\nu(a)$ vary smoothly across standardized age $a = \frac{\text{age} - \overline{\text{age}}}{\text{SD}(\text{age})}$ as polynomial functions under a log link:

$$\log \mu(a) = \sum_{k=0}^{K_\mu} \beta_k a^k, \qquad \log \nu(a) = \sum_{k=0}^{K_\nu} \gamma_k a^k$$

Parameters are estimated simultaneously by maximum likelihood using L-BFGS-B.


# Modeling Example

We demonstrate CMP modeling using a representative speed cognitive cancellation test (`speed`). In this test, children and adolescents work under a strict time limit; the raw score is the number of correctly identified targets.

```r
library(cNORM)

# Inspect dataset
head(speed)
```

## Basic Model Fitting

Fitting a CMP continuous norming model is straightforward using `cnorm.cmp()`. By default, the function uses a cubic polynomial for location (`mu_degree = 3`) and a quadratic trend for dispersion (`nu_degree = 2`):

```{r fig1, fig.height = 4, fig.width = 7}
# Fit CMP model (automatically displays diagnostic percentile plot)
model.cmp <- cnorm.cmp(
  age = speed$age,
  score = speed$fluency,
  mu_degree = 3,
  nu_degree = 2,
  plot = TRUE
)
```

Notice that because count scores are discrete integers, the model-implied percentile curves are rendered as step functions, reflecting the probability mass of integer scores.

## Model Diagnostics and Summary

Comprehensive model fit statistics, information criteria, and coefficient significance tests are available via `summary()` and `diagnostics.cmp()`:

```{r}
# Display model summary
summary(model.cmp, age = speed$age, score = speed$fluency)
```

The output confirms:
- **Location parameters ($\log \mu$):** Reflect the developmental trajectory of performance pacing over age.
- **Dispersion parameters ($\log \nu$):** Positive values indicate $\nu > 1$, confirming significant under-dispersion ($\text{Var} < \text{Mean}$) in this speeded test.


# Automatic Model Selection via BIC

To determine the optimal balance between developmental curvature and model parsimony, `cNORM` provides `autoselect.cmp()`. It evaluates an exhaustive grid of polynomial combinations across $\mu$ and $\nu$ and selects the candidate that minimizes the Bayesian Information Criterion (BIC):

```{r, eval = FALSE}
# Grid search over mu (degrees 1:4) and nu (degrees 0:2; -1 = fixed Poisson)
model.best <- autoselect.cmp(
  age = speed$age,
  score = speed$fluency,
  max_mu = 4,
  max_nu = 2,
  min_nu = 0,
  parallel = TRUE
)

# Inspect evaluated models sorted by BIC
head(model.best$selection$evaluated)
```


# Norm Score Generation

Norm scores can be predicted for individual cases or tabulated as comprehensive age-specific norm tables.

## Individual Predictions (Mid-p Adjusted)

Because raw scores are discrete counts, assigning percentiles based on $P(Y \le x)$ creates a conservative bias for tied scores. `cNORM` applies the **mid-p percentile rank convention** by default:

$$\text{PR}(x) = 100 \cdot \left[ P(Y < x) + \frac{1}{2} P(Y = x) \right]$$

```{r}
# Predict norm scores (T-scale: Mean = 50, SD = 10) for specific age-score combinations
cases <- data.frame(
  age = c(7.5, 8.5, 9.5, 10.5),
  raw = c(18, 24, 30, 35)
)

cases$T_Score <- predict(model.cmp, age = cases$age, score = cases$raw)
cases$PR <- predict(model.cmp, age = cases$age, score = cases$raw, range = Inf) # Mid-p PR
cases
```

## Discrete Norm Tables

Norm tables are generated via `normTable()` or `normTable.cmp()`. When a reliability coefficient is specified, confidence intervals based on Kelley's true score estimation (regression to the mean correction) are computed:

```{r}
# Generate norm tables for ages 8.0 and 9.0
tables <- normTable(
  c(8.0, 9.0),
  model.cmp,
  reliability = 0.88,
  CI = 0.90
)

# Display head of norm table for age 8.0
head(tables[["8"]], 10)
```

The resulting table provides:
- **x**: Raw integer score
- **Px**: Point probability $P(Y = x)$
- **Pcum**: Cumulative probability $P(Y \le x)$
- **Percentile**: Mid-p adjusted percentile rank
- **z**: Standardized z-score
- **norm**: Fitted norm score (e.g., T-score)
- **lowerCI / upperCI**: True-score confidence interval limits


# Model-Implied Distributional Moments

To analyze developmental trajectories of performance spread, skewness, and kurtosis, `predictMoments()` evaluates the exact moments via discrete PMF summation:

```{r}
# Compute model-implied population moments across ages
moments <- predictMoments(model.cmp, age = seq(7, 11, by = 1), censor = FALSE)
moments
```

Notice that `variance` is strictly smaller than `mean` across all ages, directly capturing the under-dispersed pacing of the cognitive cancellation task.


# Visual Model Comparison

You can visually contrast the fitted CMP model against alternative continuous norming models (such as distribution-free Taylor polynomials or fixed Poisson models) using `compare()`:

```{r fig2, fig.height = 4, fig.width = 7}
# Fit a traditional distribution-free Taylor model for comparison
model.taylor <- cnorm(raw = speed$fluency, age = speed$age, plot = FALSE)

# Compare models visually and print goodness-of-fit comparison table
compare(
  model.taylor,
  model.cmp,
  age = speed$age,
  score = speed$fluency,
  title = "Model Comparison: Taylor Polynomial vs. Conway-Maxwell-Poisson"
)
```


# Decision Framework for Model Selection

`cNORM` provides four distinct modeling paradigms: one semi-parametric, distribution-free approach and three parametric families. Each model possesses distinct mathematical properties tailored to specific types of test data and measurement designs.

---

### Quick Selection Matrix

| Criterion | Distribution-Free (Taylor) | Beta-Binomial | Conway-Maxwell-Poisson (CMP) | Sinh-Arcsinh (SHASH) |
| :--- | :--- | :--- | :--- | :--- |
| **Score Scale** | Continuous or discrete | Discrete integer counts ($0 \le y \le n$) | Discrete integer counts ($y \ge 0$) | Continuous real numbers ($y \in \mathbb{R}$) |
| **Test Paradigm** | General ability / achievement | Power / accuracy tests | Speeded tests / rate tasks | Continuous scales, physical / RT |
| **Ceiling Bound** | Post-hoc clipping $[minRaw, maxRaw]$ | Natural upper bound ($n$ items) | Open-ended ($[0, \infty)$) | Natural infinite support ($(-\infty, \infty)$) |
| **Distributional Shape** | Any (empirical rank projection) | Binomial with Beta-mixing | Over-, equi-, or **under-dispersed** | Skewed with heavy or light tails |
| **Key Advantage** | Maximum shape flexibility | Respects fixed item bounds | Captures under-dispersion ($\text{Var} < \text{Mean}$) | Handles negative and uncentered scores |
| **Primary Function** | `cnorm()` | `cnorm.betabinomial()` | `cnorm.cmp()` | `cnorm.shash()` |

---

### Detailed Selection Guidelines

#### 1. Use Distribution-Free Models (Taylor Polynomials) when:
* **Maximum flexibility is paramount:** The data exhibit non-standard features (e.g., multimodality, local irregularities, or empirical anomalies) that cannot be captured by standard parametric families.
* **No distributional assumptions can be justified:** You prefer empirical ranking over specifying a functional density.
* **Complex interaction surfaces:** The relationship between person location and age changes irregularly across development.
* **Caution:** Because bivariate polynomials can occasionally invert in extreme regions, always inspect consistency (`checkConsistency()`) and percentile curves (`plotPercentiles()`).

#### 2. Use Beta-Binomial Models (`cnorm.betabinomial`) when:
* **The test is an unspeeded power/accuracy test:** Raw scores represent the sum of $n$ dichotomously scored items (1PL IRT / Rasch-scaled instruments).
* **There is a strict, known ceiling:** Scores cannot exceed the maximum number of items $n$.
* **Floor or ceiling effects are present:** The distribution naturally compresses near $0$ or $n$ without requiring artificial clipping.
* **Item count is small to moderate:** The discrete nature of individual item increments is psychometrically meaningful.

#### 3. Use Conway-Maxwell-Poisson Models (`cnorm.cmp`) when:
* **The test is a speeded test:** The raw score is the number of correctly processed items within a fixed time limit (e.g., cancellation tasks, coding, symbol search, reading fluency).
* **The scale is open-ended:** There is no hard item ceiling, or the ceiling is practically unattainable within the time limit.
* **The data exhibit under-dispersion ($\text{Var} < \text{Mean}$):** Speeded tasks with easy items routinely produce performance pacing with variance substantially smaller than the mean—a feature that neither Poisson nor Negative Binomial models can accommodate.
* **Counts can also be over-dispersed ($\text{Var} > \text{Mean}$):** The CMP dispersion parameter $\nu(a)$ smoothly models over- ($\nu < 1$), equi- ($\nu = 1$), and under-dispersion ($\nu > 1$) across age.

#### 4. Use Sinh-Arcsinh Models (`cnorm.shash`) when:
* **Raw scores are continuous:** Scores have decimal precision, or stem from latent scale transformations, response times, or physical measurements (e.g., height, weight, force).
* **Zero or negative values occur:** Scores are difference scores, change scores, or unstandardized index points spanning zero and negative numbers.
* **Asymmetry and tail heaviness vary smoothly across age:** The distribution shifts from Gaussian to skewed or heavy/light-tailed across developmental stages.
* **Parametric parsimony is desired:** Four smooth polynomial trajectories ($\mu, \sigma, \epsilon, \delta$) describe the complete conditional distribution surface.


# References

* Conway, R. W., & Maxwell, W. L. (1962). A queuing model with state dependent service rates. *Journal of Industrial Engineering*, 12, 132–136.
* Huang, A. (2017). Mean-parametrized Conway-Maxwell-Poisson regression models for dispersed counts. *Statistical Modelling*, 17(6), 359–380.
* Forthmann, B., Lenhard, W., Lenhard, A., & Förster, N. (2024). Flexible Item Response Modeling for Timed Reading Comprehension Assessment. The Journal of Experimental Education, 1–17. https://doi.org/10.1080/00220973.2024.2367162
* Lenhard, A., Lenhard, W., Suggate, S. & Segerer, R. (2016). A continuous solution to the norming problem. *Assessment*, 25(1), 112–125. https://doi.org/10.1177/1073191116656437
* Sellers, K. F., & Shmueli, G. (2010). A flexible regression model for count data. *The Annals of Applied Statistics*, 4(2), 943–961.
* Shmueli, G., Minka, T. P., Kadane, J. B., Borle, S., & Boatwright, P. (2005). A useful distribution for fitting discrete data: revival of the Conway-Maxwell-Poisson distribution. *Journal of the Royal Statistical Society C*, 54(1), 127–142.
