
---
title: "Modelling Norms with the Sinh-Arcsinh (shash) Distribution"
author: "Wolfgang Lenhard & Alexandra Lenhard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#output: 
#  pdf_document:
#    latex_engine: xelatex
#    toc: true  # Optional table of contents
#    number_sections: true  # Optional section numbering
vignette: >
  %\VignetteIndexEntry{Modelling Norms with the Sinh-Arcsinh (shash) Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),
  collapse = TRUE,
  comment = "#>"
)
options(width = 60)

#devtools::load_all(".")
library(cNORM)
```

The Sinh-Arcsinh (shash) distribution is very flexible and can adjust to a wide range of distributional shapes. This makes it particularly useful for modeling psychometric data, which often exhibit non-normal characteristics such as skewness and heavy tails. Unlike discrete distributions (e.g., beta-binomial), shash naturally handles continuous scores, decimal values, and unbounded measures. It can accommodate scores of zero and negative values without transformation, unlike Box-Cox approaches that require strictly positive data. Finally, skewness (ε) and tail weight (δ) can be controlled independently, allowing precise modeling of distributional characteristics. This makes it possible to model complex developmental patterns where skewness and tail weight change with age.

# Mathematical Foundation
The Sinh-Arcsinh (shash; Jones & Pewsey, 2009) distribution is defined by a transformation of a standard normal variable. If $Y$ follows a standard normal distribution, $Y \sim N(0,1)$, then the shash-distributed variable $X$ is generated by:

$$X = \mu + \sigma \cdot \sinh\left(\frac{\text{arcsinh}(Y) - \epsilon}{\delta}\right)$$

This transformation allows the resulting variable $X$ to have its location, scale, skewness, and tail weight controlled by the four parameters:

-   **μ** (mu): location parameter, shifts the distribution horizontally (similar to mean)
-   **σ** (sigma): scale parameter ($\sigma > 0$), controls the spread of the distribution (similar to standard deviation)
-   **ε** (epsilon): skewness parameter ($\epsilon = 0$ for symmetry, $\epsilon > 0$ for right skew, $\epsilon < 0$ for left skew)
-   **δ** (delta): tail weight parameter ($\delta = 1$ produces normal-like tails, $\delta > 1$ produces heavier tails, $\delta < 1$ produces lighter tails)

The probability density function involves:

$$f(x|\mu,\sigma,\epsilon,\delta) = \frac{\delta}{\sigma\sqrt{2\pi}} \cdot \frac{\cosh(\delta \cdot \text{arcsinh}(z) + \epsilon)}{\sqrt{1 + z^2}} \cdot \exp\left(-\frac{1}{2}[\sinh(\delta \cdot \text{arcsinh}(z) + \epsilon)]^2\right)$$

where $z = \frac{x - \mu}{\sigma}$. The cumulative distribution function (CDF) does not have a closed-form expression but can be computed numerically. For a given value $x$, the CDF is:

$$F(x|\mu,\sigma,\epsilon,\delta) = P(X \leq x) = \Phi[\sinh(\delta \cdot \text{arcsinh}(z) + \epsilon)]$$

where $\Phi$ is the standard normal CDF and $z = (x - \mu)/\sigma$. The quantile function (inverse CDF) can be expressed as:

$$Q(p|\mu,\sigma,\epsilon,\delta) = \mu + \sigma \cdot \sinh\left(\frac{\text{arcsinh}(\Phi^{-1}(p)) - \epsilon}{\delta}\right)$$

## Modeling the Sinh-Arcsinh over Age

In cNORM's shash implementation, we model the four parameters as polynomial functions of standardized age with mu_degree = 3, sigma_degree = 2, epsilon_degree = 2 and delta_degree = 1 as the default settings. It is advisable to keep the delta_degree parameter low, for example 1 or 2, to avoid overfitting. The tail weight parameter δ can as well be held constant by setting the 'delta_degree' parameter to NULL. In this case, a fixed delta (default = 1) will be used to reflect population characteristics, e. g. by increasing it to values delta > 1 for heterogenous samples or delta < 1 for homogenuous samples. Age is standardized as: $\text{age}_{std} = \frac{\text{age} - \overline{\text{age}}}{\text{SD}(\text{age})}$ for numerical stability during optimization. The parameters are estimated using maximum likelihood estimation.


## Application and Prerequisites

The shash distribution is particularly well-suited for continuous performance measures (reaction times, achievement scores with decimal precision), tests with floor or ceiling effects where skewness varies across age groups, heterogeneous populations with varying degrees of individual differences, change or difference scores that may include negative values, adaptive tests or measures with variable stopping rules and large-scale assessments, where distributional assumptions are complex. And of course, it is very suited for developmental studies, since the distribution shape can change systematically with age.

When applied to normative data, there are some prerequisites and considerations to ensure optimal model performance and valid results. The test scores should exhibit systematic (though not necessarily monotonic) development across the predictor variable. The shash model can capture complex non-linear relationships through polynomial fitting of the distribution parameters. The norming sample should include a sufficient sample size. It is usually much lower than in conventional norming. Nonetheless, as a rule of thumb, we recommend a minimum of 100 cases per major age group, with larger samples needed for higher polynomial degrees or complex developmental patterns. Despite empirical validation of this rule is ongoing, these sample sizes have proven effective in previous norming studies (Lenhard et al., 2019).

While shash handles continuous scores optimally, it can also model discrete scores effectively, especially when the number of possible values is large (e. g. value range > 20). As always, representativeness of the sample is important. If the sample deviates from the target population, though post-stratification weighting can help address moderate deviations from representativeness.


## Model Selection Considerations

It is advisable to explore the norm data sample prior to modeling. The polynomial degrees for the four parameters should be selected based on theoretical expectations about developmental trajectories, sample size, and model comparison criteria (AIC, BIC, cross-validation). Visual inspection of fitted curves is essential to ensure realistic patterns. Overly complex models may lead to overfitting, especially with limited data. Thus, choose polynomial degrees based on:

- Theoretical expectations about developmental trajectories
- Sample size (higher degrees require more data to avoid overfitting)
- Model comparison criteria (AIC, BIC, cross-validation)
- Visual inspection of fitted curves for realistic patterns

When adjusting the δ parameter, you can either keep it constant or model it as a polynomial function of age using the delta_degree parameter. If the delta_degree parameter is used, keep it low (1 or 2, default = 1) to avoid overfitting. In that case, the default δ is not used. To explicitely keep delta constant, please set 'delta_degree = NULL'. The delta parameter should reflect population characteristics:

- δ = 0.7-0.9: Homogeneous populations, selected samples
- δ = 1.0: General population samples with normal-like variability
- δ = 1.2-2.0: Heterogeneous populations with high individual differences


# Modeling Example

We demonstrate Sinh-Arcsinh (shash) modeling using the PPVT-4 vocabulary development dataset. Please have a look at the beta-binomial vignette for a comparison with discrete models.

## Basic Model Fitting

```{r fig1, fig.height = 4, fig.width = 7}
# Fit shash model with custom settings
# The function automatically displays percentile plots
model.shash <- cnorm.shash(age = ppvt$age, score = ppvt$raw)


# Use print(model.shash), diagnostics(model.shash) or summary(model.shash)
# to retrieve information on the data fit.
```

The model uses reasonable default polynomial degrees per default and provides immediate visual feedback through percentile plots. The output includes fitted parameters, convergence information, and basic model statistics. Model diagnostics (summary, diagnostics) provides:

- Model fit statistics (log-likelihood, AIC, BIC, R²)
- Parameter estimates with standard errors and significance tests
- Convergence diagnostics
- Separate tables for location, scale, and skewness parameters

Please pay attention to convergence (should be successful, if not, please inspect model visually), R² (correlation between fitted and manifest percentiles; > 0.95 desirable) and general fit statistics.


## Custom Model Specifications

For datasets with complex patterns, adjust polynomial degrees and distributional parameters:

```{r, eval=FALSE}
# Conservative parameterization with fixed delta across age
model.simple <- cnorm.shash(age = ppvt$age, score = ppvt$raw,
                           mu_degree = 2,        # Quadratic location pattern
                           sigma_degree = 1,     # Linear variability change  
                           epsilon_degree = 1,   # Linear skewness change
                           delta_degree = NULL,  # deactivates polynimial fitting for delta
                           delta = 1.1)          # Slightly heavy tails,
                                                 # kept constant across age

# Example with more complex parameterization
model.complex <- cnorm.shash(age = ppvt$age, score = ppvt$raw,
                           mu_degree = 4,        # Quadric pattern
                           sigma_degree = 3,     # Complex variability changes  
                           epsilon_degree = 2,   # Quadratic age-varying skewness
                           delta_degree = 2)     # Changing tail weights across age (quadratic)

# Compare models
compare(model.simple, model.complex, age = ppvt$age, score = ppvt$raw,
        title = "ShaSh Model Comparison")


```

**Note**: Higher polynomial degrees increase model flexibility but may lead to overfitting with insufficient data. Always validate complex models through visual inspection and preferable cross-validation.


## Post-Stratification and Weighting

Like betabinomial and distribution free models, cNORM supports post stratification in shash distributions via iterative post stratification to approximate representativity:

```{r, eval=FALSE}
# Calculate post-stratification weights
margins <- data.frame(variables = c("sex", "sex", "migration", "migration"),
                     levels = c(1, 2, 0, 1),
                     share = c(.52, .48, .7, .3))

weights <- computeWeights(ppvt, margins)

# Fit weighted ShaSh model
model.weighted <- cnorm.shash(ppvt$age, ppvt$raw, weights = weights)

# Compare weighted vs. unweighted
compare(model.shash, model.weighted, age = ppvt$age, score = ppvt$raw,
        title = "Unweighted vs. Weighted ShaSh Models")
```


# Norm Score Generation

Norm scores can be calculated for individual cases (or vectors of cases) or as comprehensive norm tables for specified ages.

```{r eval=FALSE}
# Individual Norm Score Prediction:
# Generate norm scores for specific age-score combinations
ages <- c(10.25, 10.75, 11.25, 11.75)
raw_scores <- c(180, 185, 190, 195)

norm_scores <- predict(model.shash, ages, raw_scores)
prediction_table <- data.frame(
  Age = ages, 
  Raw_Score = raw_scores, 
  Norm_Score = round(norm_scores, 1)
)
print(prediction_table)


# Norm Score tables:
# Generate detailed norm tables for multiple ages
tables <- normTable.shash(model.shash, 
                         ages = c(10.25, 10.75), 
                         start = 150, 
                         end = 220,
                         step = 1,
                         CI = 0.95, 
                         reliability = 0.94)

# Display head from first table
head(tables[[1]], 10)
```

The norm tables provide:

- **x**: Raw scores
- **Px**: Probability density values
- **Pcum**: Cumulative probabilities
- **Percentile**: Percentile ranks (0-100)
- **z**: Standardized z-scores  
- **norm**: Norm scores in specified scale
- **Confidence intervals**: When reliability is specified


# Model Comparison and Selection

## Comparing Distributional Approaches

```{r fig5, fig.height = 4, fig.width = 7}
# Compare shash with betabinomial models (BB). BB models should work worse,
# since the test has stop rules, leading to non-binomial distributions.
model.bb <- cnorm.betabinomial(ppvt$age, ppvt$raw, n = 228, plot = FALSE)

# Model comparisons
compare(model.shash, model.bb, age = ppvt$group, score = ppvt$raw,
        title = "SinH-ArcSinH vs. Beta-Binomial")


# Compare distribution free Taylor model
model.taylor <- cnorm(group = ppvt$group, raw = ppvt$raw, plot=FALSE)

# Model comparisons shash versus taylor
compare(model.shash, model.taylor, age = ppvt$group, score = ppvt$raw,
        title = "SinH-ArcSinH vs. Taylor")
```

## Decision Framework for Model Selection

`cNORM` provides four distinct modeling paradigms: one semi-parametric, distribution-free approach and three parametric families. Each model possesses distinct mathematical properties tailored to specific types of test data and measurement designs.

---

### Quick Selection Matrix

| Criterion | Distribution-Free (Taylor) | Beta-Binomial | Conway-Maxwell-Poisson (CMP) | Sinh-Arcsinh (SHASH) |
| :--- | :--- | :--- | :--- | :--- |
| **Score Scale** | Continuous or discrete | Discrete integer counts ($0 \le y \le n$) | Discrete integer counts ($y \ge 0$) | Continuous real numbers ($y \in \mathbb{R}$) |
| **Test Paradigm** | General ability / achievement | Power / accuracy tests | Speeded tests / rate tasks | Continuous scales, physical / RT |
| **Ceiling Bound** | Post-hoc clipping $[minRaw, maxRaw]$ | Natural upper bound ($n$ items) | Open-ended ($[0, \infty)$) | Natural infinite support ($(-\infty, \infty)$) |
| **Distributional Shape** | Any (empirical rank projection) | Binomial with Beta-mixing | Over-, equi-, or **under-dispersed** | Skewed with heavy or light tails |
| **Key Advantage** | Maximum shape flexibility | Respects fixed item bounds | Captures under-dispersion ($\text{Var} < \text{Mean}$) | Handles negative and uncentered scores |
| **Primary Function** | `cnorm()` | `cnorm.betabinomial()` | `cnorm.cmp()` | `cnorm.shash()` |

---

### Detailed Selection Guidelines

#### 1. Use Distribution-Free Models (Taylor Polynomials) when:
* **Maximum flexibility is paramount:** The data exhibit non-standard features (e.g., multimodality, local irregularities, or empirical anomalies) that cannot be captured by standard parametric families.
* **No distributional assumptions can be justified:** You prefer empirical ranking over specifying a functional density.
* **Complex interaction surfaces:** The relationship between person location and age changes irregularly across development.
* **Caution:** Because bivariate polynomials can occasionally invert in extreme regions, always inspect consistency (`checkConsistency()`) and percentile curves (`plotPercentiles()`).

#### 2. Use Beta-Binomial Models (`cnorm.betabinomial`) when:
* **The test is an unspeeded power/accuracy test:** Raw scores represent the sum of $n$ dichotomously scored items (1PL IRT / Rasch-scaled instruments).
* **There is a strict, known ceiling:** Scores cannot exceed the maximum number of items $n$.
* **Floor or ceiling effects are present:** The distribution naturally compresses near $0$ or $n$ without requiring artificial clipping.
* **Item count is small to moderate:** The discrete nature of individual item increments is psychometrically meaningful.

#### 3. Use Conway-Maxwell-Poisson Models (`cnorm.cmp`) when:
* **The test is a speeded test:** The raw score is the number of correctly processed items within a fixed time limit (e.g., cancellation tasks, coding, symbol search, reading fluency).
* **The scale is open-ended:** There is no hard item ceiling, or the ceiling is practically unattainable within the time limit.
* **The data exhibit under-dispersion ($\text{Var} < \text{Mean}$):** Speeded tasks with easy items routinely produce performance pacing with variance substantially smaller than the mean—a feature that neither Poisson nor Negative Binomial models can accommodate.
* **Counts can also be over-dispersed ($\text{Var} > \text{Mean}$):** The CMP dispersion parameter $\nu(a)$ smoothly models over- ($\nu < 1$), equi- ($\nu = 1$), and under-dispersion ($\nu > 1$) across age.

#### 4. Use Sinh-Arcsinh Models (`cnorm.shash`) when:
* **Raw scores are continuous:** Scores have decimal precision, or stem from latent scale transformations, response times, or physical measurements (e.g., height, weight, force).
* **Zero or negative values occur:** Scores are difference scores, change scores, or unstandardized index points spanning zero and negative numbers.
* **Asymmetry and tail heaviness vary smoothly across age:** The distribution shifts from Gaussian to skewed or heavy/light-tailed across developmental stages.
* **Parametric parsimony is desired:** Four smooth polynomial trajectories ($\mu, \sigma, \epsilon, \delta$) describe the complete conditional distribution surface.

---

### Empirical Model Evaluation

When choosing between candidate models or tuning polynomial degrees, evaluate models across multiple complementary criteria:

1. **Theoretical Appropriateness:** Does the model's support and dispersion mechanism match the test design (e.g., Beta-Binomial for accuracy tests, CMP for speeded tests)?
2. **Information Criteria (AIC & BIC):** Lower values indicate better trade-offs between goodness-of-fit and parsimony. (Use `autoselect.cmp()`, `autoselect.betabinomial()`, or `autoselect.shash()` for automated grid selection).
3. **Cross-Validation (`cnorm.cv`):** Evaluate out-of-sample prediction accuracy and check crossfit indices ($\approx 1.0$) to guard against overfitting.
4. **Direct Model Comparison (`compare()`):** Overlay percentile curves from competing models (e.g., Taylor vs. CMP) and inspect differences in $R^2$, RMSE, and bias against manifest group percentiles.
5. **Visual Plausibility:** Percentile curves must progress smoothly across age without crossings, implausible compressions, or edge-extrapolation artifacts.


# Conclusion

The Sinh-Arcsinh (shash) distribution provides a sophisticated yet practical framework for modeling psychometric norm data. Its ability to independently control location, scale, skewness, and tail weight makes it particularly valuable for applications where traditional parametric approaches are inadequate due to distributional complexity. Best practices for shash modeling:

1. Carefully assess data characteristics before model selection
2. Use appropriate sample sizes for model complexity
3. Validate models through visually and via statistical indicators (R2, AIC, Bias)
4. Compare alternative distributional assumptions. Is distribution-free or beta-binomial more appropriate?
5. Document model selection rationale


# References

Jones, M. C., & Pewsey, A. (2009). Sinh-arcsinh distributions. *Biometrika*, 96(4), 761-780.

Lenhard, A., Lenhard, W., Gary, S. (2019). Continuous norming of psychometric tests: A simulation study of parametric and semi-parametric approaches. *PLoS ONE*, 14(9), e0222279. https://doi.org/10.1371/journal.pone.0222279
